Gravitational wave sources in our Galactic backyard: LISA predictions for BH and NS binaries

### 2021-06-02 paper/tex/sections/methods.tex
Updates from Overleaf

Most previous studies that predict a detection rate for LISA place binaries in the Milky Way independently of their age or evolution. We improve upon this as the first study to use an empirically-informed analytical model of the Milky Way that takes into account the galaxy's enrichment history by applying the metallicity-radius-time relation from \citet{Frankel+2018}. The authors developed this relation in order to measure the global efficiency of radial migration in the Milky Way and calibrated it using a sample of red clump stars measured with APOGEE \citep{Majewski+2017} and matched to stars in Gaia \citep{GaiaCollaboration+2016}.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{1_galaxy_diagram.png}
    \caption{A schematic illustrating how we create a mock Milky Way galaxy. On the left, we show the various individual distributions for the different parameters. In the star formation history distribution, we also plot the separate distributions of each component with dotted lines to more clearly explain the shape. On the right, we show an example instance of the Milky Way with $10^5$ binaries shown as points, where the colour of each point represents its metallicity. The top panel shows a side-on view and the bottom panel shows a view from above.}
    \label{fig:galaxy_schematic}
\end{figure*}

Our model for the Milky Way accounts for the thin disc, thick disc and central bar/bulge. For each of these components we use a separate star formation history, radial and vertical distribution, which we combine into a single model, weighting each component by its stellar mass. \citet{Licquia+2015} gives that the stellar mass of the bulge is $0.9 \times 10^{10} \unit{M_{\odot}}$ and the stellar mass of the disc is $5.2 \times 10^{10} \unit{M_\odot}$, which we split equally between the thin and thick disc. \todo{@Neige, do you have a suggestion for a citation for having equal mass thin/thick discs}.

\textit{Star formation history:} For the thin and thick discs we use the star formation history given by \citet[][Eq.~2]{Frankel+2018},
\begin{equation}\label{eq:thin_disc_tau}
    p(\tau) \propto \exp \qty(-\frac{(\tau_m - \tau)}{\tau_{\rm SFR}}),
\end{equation}
where $\tau$ is the lookback time (the amount of time elapsed between the binary's zero-age main sequence and today), $\tau_m = 12 \unit{Gyr}$ is the assumed age of the Milky Way and $\tau_{\rm SFR} = 6.8 \unit{Gyr}$ is the star formation timescale. Following \citet{Frankel+2018}, we assume that there is a transition from thick disc star formation to thin disc star formation $8 \unit{Gyr}$ ago and apply a sharp transition at this point. Despite its simplicity, this has been shown to reproduce the Milky Way age and abundance distributions well \citep{Sharma+2020} \todo{@Neige, is this justification correct/relaying the results of Sharma+2020 properly?}. Therefore, for the thin disc we normalise this distribution on the range $[0, 8] \unit{Gyr}$ and for the thick disc we normalise on the range $[8, 12] \unit{Gyr}$.

For the bulge, we base our distribution on the one presented in \citet[][Fig.~7]{Bovy+2019}. This model is fit using the data from APOGEE and Gaia as \citet{Frankel+2018}. \todo{@Neige, I need a sentence here to explain what biases we account for and so why we shift it. I recall it's something to do with fainter stars being missed and the bright ones not being part of the bulge but I can't articulate it very well, could you please help me with this?}. We therefore approximate the distribution using a $\beta(2,3)$ distribution, shifted and scaled such that stars are only formed in the range $[6, 12] \unit{Gyr}$.

\textit{Radial distribution:} For each of the three components we employ the same single exponential distribution (but with different scale lengths)
\begin{equation}\label{eq:galaxy_R}
    p(R) = \exp(-\frac{R}{R_d}) \frac{R}{R_d^2},
\end{equation}
where $R$ is the Galactocentric radius and $R_d$ is the scale length of the component. For the thin disc, we set $R_d = R_{\rm exp}(\tau)$, where $R_{\rm exp}(\tau)$ is the scale length presented in \citet[][Eq.~5]{Frankel+2018}
\begin{equation}
    R_{\rm exp}(\tau) = 4 \unit{kpc} \qty(1 - \alpha_{R_{\rm exp}} \qty(\frac{\tau}{8 \unit{Gyr}})),
\end{equation}
where $\alpha_{R_{\rm exp}} = 0.3$ is the inside-out growth parameter\footnote{In $R_{\rm exp}(\tau)$, we use 4 kpc instead of 3 kpc for the 0 Gyr exponential scale-length of the disc as NF finds that it provides a better fit to the original data}. This scale length accounts for the inside-out growth of the thin disc and hence is age dependent. We assume $R_d = (1 / 0.43) \unit{kpc}$ for the thick disc \citep[][Table~1]{Bovy+2016} and $R_d = 1.5 \unit{kpc}$ for the bulge \citep{Bovy+2019}.

\textit{Vertical distribution}: Similar to the radial distribution, we use the same single exponential distribution (but with different scale heights) for each component
\begin{equation}\label{eq:galaxy_z}
    p(\abs{z}) = \frac{1}{z_d} \exp\qty(-\frac{z}{z_d}),
\end{equation}
where $z$ is the height above the Galactic plane and $z_d$ is the scale height. We set $z_d = 0.3 \unit{kpc}$ for the thin disc \citep{McMillan+2011} and $z_d = 0.95 \unit{kpc}$ for the thick disc \citep{Bovy+2016}. For the bulge, we set $z_d = 1.5 \unit{kpc}$ such that the probability of forming a star approaches zero at approximately the co-rotation radius of the Galactic bar/bulge \citep{Bovy+2019}. \todo{@Neige, did I justify the bulge/bar height correctly?}

@@

Although \citet{Frankel+2018} only fit this model for the thin disc, we use this same metallicity-radius-time relation for each component of the galaxy. \citet{Sharma+2020} showed that using a simple continuous model for both the thin and thick disc the Milky Way abundance distributions well and hence we can apply this same relation to the thick disc. \citet{Bovy+2019} show in Fig.~7 that the abundance distribution and chemical tracks of the stars in the bulge/bar follow the same track as those outside and hence we can use the same metallicity-radius-time relation.

>>>>

Most previous studies that predict a detection rate for LISA place binaries in the Milky Way independently of their age or evolution. We improve upon this as the first study to use an empirically-informed analytical model of the Milky Way that takes into account the galaxy's enrichment history by applying the metallicity-radius-time relation from \citet{Frankel+2018}. The authors developed this relation in order to measure the global efficiency of radial migration in the Milky Way and calibrated it using a sample of red clump stars measured with APOGEE \citep{Majewski+2017}.
%and matched to stars in Gaia \citep{GaiaCollaboration+2016}.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{1_galaxy_diagram.png}
    \caption{A schematic illustrating how we create a mock Milky Way galaxy. The left panel illustrates the different model aspects: star formation history of 3 galactic components (individually shown in the dotted lines), spatial distribution at birth, age-metallicity-radius relation, and vertical distribution.
    On the right, we show an example instance of the Milky Way with $10^5$ binaries shown as points colour coded by metallicity. The top panel shows a side-on view and the bottom panel shows a face-on view.}
    \label{fig:galaxy_schematic}
\end{figure*}

Our model for the Milky Way accounts for the low-$\alpha$ disc, high-$\alpha$ disc and central bar/bulge. For each of these components, we use a separate star formation history, radial and vertical distribution, which we combine into a single model, weighting each component by its stellar mass. \citet{Licquia+2015} gives that the stellar mass of the bulge is $0.9 \times 10^{10} \unit{M_{\odot}}$ and the stellar mass of the disc is $5.2 \times 10^{10} \unit{M_\odot}$, which we split equally between the low- and high-$\alpha$ discs \citep[e.g.,][]{Snaith+2014}.

\textit{Star formation history:} 
We use an exponentially declining star formation history \citep{Frankel+2018} (a priori inspired by average the cosmic star formation history) for the combined low- and high-$\alpha$ disks, where the two disks transition at about 8 Gyr ago, and re-normalize the produced mass to be equal in each of the two components.
\begin{equation}\label{eq:thin_disc_tau}
    p(\tau) \propto \exp \qty(-\frac{(\tau_m - \tau)}{\tau_{\rm SFR}}),
\end{equation}
where $\tau$ is the lookback time (the amount of time elapsed between the binary's zero-age main sequence and today), $\tau_m = 12 \unit{Gyr}$ is the assumed age of the Milky Way and $\tau_{\rm SFR} = 6.8 \unit{Gyr}$ is the star formation timescale.

The star formation history of the Milky Way bulge (which we assume here to be dominated by the central bar) has many uncertainties due to (1) sizeable age measurement uncertainties at large ages in observational studies, (2) complex selection processes affecting the observed age distributions, and (3) formation mechanisms still under debate. But the bulge was shown to contain stars with a range of 6-12 Gyr \citep[e.g.,][]{Bovy+2019}, where the younger tail of ages might come from the growth of the Galactic bar. To model the bulge age distribution more realistically than in previous studies (assuming an older bulge coming from a single starburst), we choose to adopt a star formation history using a $\beta(2,3)$ distribution, shifted and scaled such that stars are only formed in the range $[6, 12] \unit{Gyr}$

\textit{Radial distribution:} For each of the three components we employ the same single exponential distribution (but with different scale lengths)
\begin{equation}\label{eq:galaxy_R}
    p(R) = \exp(-\frac{R}{R_d}) \frac{R}{R_d^2},
\end{equation}
where $R$ is the Galactocentric radius and $R_d$ is the scale length of the component. For the low-$\alpha$ disc, we set $R_d = R_{\rm exp}(\tau)$, where $R_{\rm exp}(\tau)$ is the scale length presented in \citet[][Eq.~5]{Frankel+2018}
\begin{equation}
    R_{\rm exp}(\tau) = 4 \unit{kpc} \qty(1 - \alpha_{R_{\rm exp}} \qty(\frac{\tau}{8 \unit{Gyr}})),
\end{equation}
where $\alpha_{R_{\rm exp}} = 0.3$ is the inside-out growth parameter\footnote{In $R_{\rm exp}(\tau)$, we use 4 kpc instead of 3 kpc for the 0 Gyr exponential scale-length of the disc as NF finds that it provides a better fit to the original data}. This scale length accounts for the inside-out growth of the low-$\alpha$ disc and hence is age dependent. We assume $R_d = (1 / 0.43) \unit{kpc}$ for the high-$\alpha$ disc \citep[][Table~1]{Bovy+2016} and $R_d = 1.5 \unit{kpc}$ for the bar component \citep{Bovy+2019}.

\textit{Vertical distribution}: Similar to the radial distribution, we use the same single exponential distribution (but with different scale heights) for each component
\begin{equation}\label{eq:galaxy_z}
    p(\abs{z}) = \frac{1}{z_d} \exp\qty(-\frac{z}{z_d}),
\end{equation}
where $z$ is the height above the Galactic plane and $z_d$ is the scale height. We set $z_d = 0.3 \unit{kpc}$ for the low-$\alpha$ disc \citep{McMillan+2011} and $z_d = 0.95 \unit{kpc}$ for the high-$\alpha$ disc \citep{Bovy+2016}. 
\todo{Neige to Tom: for bar, I will think more for your bar scale-height, but the exact value should not change your results much since they will only slightly change the distance distribution of your sources, but not the rates nor the normalization.}
For the bulge, we set $z_d = 1.5 \unit{kpc}$ such that the probability of forming a star approaches zero at approximately the co-rotation radius of the Galactic bar/bulge \citep{Bovy+2019}. \todo{@Neige, did I justify the bulge/bar height correctly? A: The vertical distribution should not influence at which radius the bar should stop forming stars -- more later}

@@

Although \citet{Frankel+2018} only fit this model for the low-$\alpha$ disc, we also use this metallicity-radius-time relation for the high-$\alpha$ disk and the bar, but focusing on the chemical tracks more representative to the inner disk and large ages. \citet{Sharma+2020} showed that using a simple continuous model for both the low- and high-$\alpha$ discs, the Milky Way abundance distributions could be well reproduced. Empirically, the chemical tracks in the [$\alpha$/Fe]-[Fe/H] plane of the stars in the bulge/bar follow the same track as those of the old stars in the Solar neighbourhood \citep[][Fig.~7,]{Bovy+2019}, which motivates our modelling choice to use the same metallicity-radius-time relation.
